Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-03-05 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Updated FDS ref and developing FDS section text

Similar to the previous chapter the goal is to utilize an analytical tool to predict or simulate the results from some of the key experiments conducted in this study. Unlike the previous chapter the computational fluid dynamics (CFD) models are not as widely used by fire investigators as the algebraic tools are.  The CFD models require more knowledge, more computer resources, and more input data to utilize correctly.  Currently modelers need to develop there own data or rely on poorly defined databases with the hope that the material tested in the database is similar to the materials of interest.

@@

The Fire Dynamics Simulator (FDS) and the scientific visualization tool, Smokeview, are the most widely used and well documented CFD tools in the fire protection engineering community. In addition these tools have on going verification and validation programs.  Given the extent and high quality of the existing FDS documentation, there is no value to attempting to recreate a description of the model here other than to direct the reader to the volumes of information and data that exist for the FDS and Smookeview suite.

FDS~\cite{FDS_Users_Guide} is a CFD model initially developed and currently maintained by NIST that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires. A CFD model has a computational domain that is divided into  three-dimensional rectangular control volumes or computational cells. The size and arrangement of computational cells is sometimes referred to as the grid or the mesh.  The CFD model computes the density, velocity, temperature, pressure, and gas concentrations in each cell. Based on the laws of conservation of mass, momentum, and energy, the model tracks the generation and movement of fire gases. One of the most important aspects of FDS is that it is mathematically verified~\cite{FDS_Verification_Guide}and validated against fire test data~\cite{FDS_Validation_Guide}to ensure that it is accurate and provides the expected results, given appropriate input data. A complete description of the FDS model is provided in the FDS Technical Reference Guide~\cite{FDS_Tech_Guide}.

Smokeview is a software tool specifically designed to visualize simulation results from FDS~\cite{Smokeview_Users_Guide}. Smokeview visualizes smoke and other attributes of the fire simulation by showing tracer particle flow and two dimensional (2D) or three dimensional (3D) shaded contours of gas flow data, including temperature and flow vectors showing flow direction and magnitude. Smokeview has the ability to visualize fire and smoke by displaying a series of partially transparent planes where the transparencies in each plane (at each grid node) are determined from soot densities computed by FDS. Smokeview can also visualize data at particular snapshots in time using 2D or 3D contours of data that show temperature, flow direction, and flow magnitude.  Smokeview also has a feature that can represent a fire or char pattern on a surface based on user supplied threshold criteria.

@@

The source fire simulation will be compared against the source fire characterization data to determine the best grid cell resolution and other inputs to use to provide the best fire simulation for use in the simulations of the wall and compartment experiments.

@@

One of the key determinations that is required when developing an FDS model is the mesh or grid cell size.  If a single plane of the data from the model were compared to a digital image we might refer to the grid cell size as the resolution of the image.  The smaller the grid cells the better the resolution. However increased resolution or the smaller grid cell size, the longer it will take the simulation to run.

@@

 The FDS documents provide a method to determine how well a flow field will be resolved buy using the non-dimensional expression ${D^*}/{\delta x}$, where $D^*$ is a characteristic fire diameter and $\delta x$ is the nominal size of a mesh or grid cell. This value has been referred to as the plume resolution index. Given that the fire plume region is the focus of this study, it is important to understand how changes in plume mesh resolution impact the results.

@@

To examime the impact of computational grid cell size on the results, three different computational cell sizes; 80~mm, 40~mm, and 20~mm were choosen to model the fire. The resulting plume resolution index (RI) values were approximately 4.4, 8.8, and 17.5 respectively. Given the range of grid cell sizes and the characteristic diameter of fire, 0.35~m, this resulted in what could be considered a course (RI=5), medium (RI=10), and fine (RI=20) plume resolution results from the grid cell size choices.

@@

>>>>

The goal of this chapter is to utilize an analytical tool to predict or simulate the results from some of the key experiments conducted in this study. Unlike the previous chapter the computational fluid dynamics (CFD) models are not as widely used by fire investigators as the algebraic tools are.  The CFD models require more knowledge, more computer resources, and more input data to utilize correctly.  Currently modelers need to develop there own data or rely on poorly defined databases with the hope that the material tested in the database is similar to the specific material of interest.

@@

The Fire Dynamics Simulator (FDS) and the scientific visualization tool, Smokeview, are the most widely used and well documented CFD tools in the fire protection engineering community. In addition, these tools have on going verification and validation programs.  Given the extent and high quality of the existing FDS documentation, there is no value to attempting to recreate a description of the model here other than to direct the reader to the volumes of information and data repository that exist for the FDS and Smookeview suite.

FDS~\cite{FDS_Users_Guide} is a CFD model initially developed and currently maintained by NIST that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires. A CFD model has a computational domain that is divided into  three-dimensional rectangular control volumes or computational cells. The size and arrangement of computational cells is sometimes referred to as the grid or the mesh.  The CFD model computes the density, velocity, temperature, pressure, and gas concentrations in each cell. Based on the laws of conservation of mass, momentum, and energy, the model tracks the generation and movement of fire gases. One of the most important aspects of FDS is that it is mathematically verified~\cite{FDS_Verification_Guide}and validated against fire test data~\cite{FDS_Validation_Guide}to ensure that it is accurate and provides the expected results, given appropriate input data.  The data that FDS has been validated against can be found on-line is repository of the reports that were the source of the data~\cite{FDS-SMV_repository}. A complete description of the FDS model is provided in the FDS Technical Reference Guide~\cite{FDS_Tech_Guide}.

Smokeview is a software tool specifically designed to visualize simulation results from FDS~\cite{Smokeview_Users_Guide}. Smokeview visualizes smoke and other attributes of the fire simulation by showing tracer particle flow and two dimensional (2D) or three dimensional (3D) shaded contours of gas flow data, including temperature and flow vectors showing flow direction and magnitude. Smokeview has the ability to visualize fire and smoke by displaying a series of partially transparent planes where the transparencies in each plane (at each grid node) are determined from soot densities computed by FDS. Smokeview can also visualize data at particular snapshots in time using 2D or 3D contours of data that show temperature, flow direction, and flow magnitude.  Smokeview also has a feature that can represent a fire or char pattern on a surface based on user supplied threshold criteria.  Smokeview also has a Technical Reference Guide~\cite{Smokeview_Tech_Guide} and a Verification Guide~\cite{Smokeview_Verification_Guide}.

@@

Several source fire simulations will be modelled with a range of grid cell resolutions and radiative fraction values.  The results are compared against the source fire characterization data to determine which grid cell resolution and radiative fraction will provide the best fire simulation for use in the simulations of the wall and compartment experiments.

@@

One of the key determinations that is required when developing an FDS model is the mesh or grid cell size.  If a single plane of the data from the model were compared to a digital image we might refer to the grid cell size as the resolution of the image.  The smaller the grid cells the better the resolution. However increased resolution will require more computational resources and additional run time. The question is does it provide a more realistic answer.  As with most pursuits there is no one answer.  The `best' grid size depends on the scenario being modeled.

@@

 The FDS User Guide provides a method to determine how well a flow field will be resolved buy using the non-dimensional expression ${D^*}/{\delta x}$, where $D^*$ is a characteristic fire diameter and $\delta x$ is the nominal size of a mesh or grid cell. This value has been referred to as the plume resolution index. Given that the fire plume region is the focus of this study, it is important to understand how changes in plume mesh resolution impact the results.

@@

To examime the impact of computational grid cell size on the results, three different computational cell sizes; 80~mm, 40~mm, and 20~mm were choosen to model the fire. The resulting mesh resolution index (RI) values were approximately 4.4, 8.8, and 17.5, respectively. Given the range of grid cell sizes and the characteristic diameter of fire, 0.35~m, this resulted in what could be considered a course (RI=5), medium (RI=10), and fine (RI=20) mesh resolution.  FDS simulations of the source fire characterization experiments will be conducted with these mesh resolutions and compared with the results from the experiments.

@@

Source term model description ....
